Network based multifactorial modelling microRNA-target interaction.

### 2019-07-22 article.Rmd
rewrite method section

---
title: "Network based multifactorial modelling microRNA-target interaction."
author: 
- Selcen Ari^[Yildiz Technical University Department of Bioengineering]
- Alper Yilmaz^[Yildiz Technical University Department of Bioengineering]
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document: 
    keep_tex: true
  html_document: default
bibliography: ref.bib
---

@@

As it is possible for miRNAs to suppress multiple targets, an individual mRNA molecule can also be targeted by multiple miRNAs. In that case, the targeted mRNAs exhibit competitor behavior, that is hypothesized as competing endogenous RNAs (ceRNAs) [@ala_integrated_2013; @cesana_deciphering_2013], against their miRNAs. Briefly, Ala et al. have explained the ceRNA hypothesis as disturbance of the other target when one of the targets on a steady-state system that included one miRNA and two target was perturbed with expression change [@ala_integrated_2013]. Regarding interaction between miRNAs and their target in a cell, explaining and predicting results of an individual perturbation is difficult due to complexity of interactions. Various computational and experimental studies have tackled the problem of unraveling ceRNA:miRNA interactions. It has been observed that when abundance of one of the targets of miR-122 was increased, the other target expression also slightly increased as a result of decreasing repressive activity of miR-122 on them [@denzler_assessing_2014]. Bosson et al. have developed a mathematical model for changes on total target pool concentration after grouping targets according to affinity and demonstrated that miRNA activity correlated with affinity between miRNA and target [@bosson_endogenous_2014]. Cooperative efficiency of miRNAs as well as competitor behaviors of targets were also studied and it has been demonstrated to be crucial for regulating available mRNA levels of targets [@denzler_impact_2016]. MiRNA:target interactions have been modeled as stoichiometric and catalytic mechanisms and Figliuzzi et al. have recommended handling models in network context [@figliuzzi_micrornas_2013]. The model that can explain miRNA target interaction through topological features has been applied at bipartite network by Nitzan et al. [@nitzan_interactions_2014]. Robinson and Henderson applied the model that handles miRNA:target direct and indirect interactions via common miRNA of gene and target of miRNA, at bipartite network. It has been demonstrated that all miRNAs and targets in the network can interact with each other through common miRNAs and genes, without interaction between the same type of nodes [@robinson_modelling_2018].Associated genes that are targets of the same miRNAs have been found with help of correlation of gene expression changes in recent algoritm [@markus_list_sponge_2017]. List et al. have specified that their approach can be useful for ceRNA studies and published their approach as an R package.

## Methods
### miRNA:target ratio theoretical action in network based ceRNA model
We developed a novel network based approach that can use to simulation of gene regulation after any perturbation at miRNA:target interactions. If a perturbation or change induces on graph, whole interacted nodes vary because of changes in miRNA:target ratios. The system was assumed as steady state that miRNAs affect according to the ratio of the individual target to all targets (\autoref{fig1}a, expression values shown as integer for comprehensible sampling). When expression of gene 2 (G2) increased to two fold (\autoref{fig1}b), distribution of miRNA (M1) expression changes (\autoref{fig1}c). Consequently, expression of common target (G4) of miRNA1 and miRNA2 also varies. The change at expression of common target triggers to changes of proportional distribution of miRNA2 (\autoref{fig1}d).

![Schematic presentation of mechanism of network based model. a) In steady state, miRNAs (M: gray triangle) repress targets (G: circle coconut) according to proportion of their targets' expression. b-c-d) MiRNA expression distributions to targets vary when expression of one of the targets perturbed.](Fig1.jpg){#fig1}

The model was developed by making some assumptions. Transcription rates and degradation rates of miRNAs were assumed as stationary and equal. Targets also have stable transcription and degradation rates, at the same time they are equal. The factors that could be affect the miRNA:target interaction are disregarded in this approach. It is considered that miRNAs degrades or inhibits them via any mechanisms after binding to target while the miRNAs were used reversibly. In initial conditions, gene expression amounts in system have been considered after repression activity of miRNAs. If a change occurs in the system, variation on miRNA proportional distribution are reflected as change of repression activity.
\begin{equation} 
    Eff_{gi}= C_m * C_{gi}/\sum_{1}^{i} C_g \tag{1}\label{eq:1}
\end{equation}
\begin{equation}
   R_{gi}= Eff_{i1}+ Eff_{i2} ... \tag{2}\label{eq:2}
\end{equation}

The efficiency of a miRNA on the individual target ($Eff_{gi}$) is calculated according to equation \eqref{eq:1}; $C_m$, miRNA expression in the group; $C_{gi}$, symbolize individual gene expression; $C_g$ gene expressions in the group (Groups are determined depending on miRNAs. Targets that are interacted with the same miRNA find in a group.). In this approach, because a gene can be found in different groups, the cooperative activity of microRNAs also can be calculated. Total repression activity of different miRNAs on the common target, $R$, is calculated based on equation \eqref{eq:2}; sum of efficiency of different miRNAs on the target.

![Calculations to determine of miRNA binding and repression efficiency. $G$, Gene; $M$, miRNA; $STE$, seed type effect; $RE$, Region Effect; $E$, Energy; $STE’$, normalized values of seed type efficiency coefficient; $RE’$, normalized values of region efficiency coefficient; $E’$, normalized values of energy coefficient.](fig2.jpg){#fig2}

### miRNA action in network based multifactorial ceRNA model

Interactions between miRNAs and their targets can be affected from different parameters. For this reason, we improved model with factors that can affect to miRNA activity. In our approach, the parameters are classified as binding and efficiency factors and all of these are evaluated in their groups. Binding factors determine interaction between miRNA and target. The efficiency factors decide functionality of binding. In the literature, binding free energy [@cao_predicting_2012 ; @helwak_mapping_2013] and seed type [@werfel_preferential_2017] in miRNA:target interactions are represented as binding affinity. Efficiency factors determine how many of microRNA:target complexes will result in inhibition. For example, binding region on the target shows whether the target can be inhibited (functionality)[@hausser_analysis_2013; @helwak_mapping_2013]. Both of binding and efficiency factors are evaluated in the groups and normalized according to equation \eqref{eq:3}; each factor is normalized based on maximum variables that found in their groups.
\begin{equation}
F`= F/F_{max} \tag{3}\label{eq:3}
\end{equation}

The normalized values of factors take into account to determine binding activity and miRNA efficiency on targets (\autoref{fig2}). Binding affinities (activity, $Eff$) of miRNAs on each individual gene are calculated as shown in equation \eqref{eq:4}; $C_m$, miRNA expression in the group; $C_g$, Gene expression; $C_{gi}$, individual gene expression; gi, individual gene; g, whole genes in a group (\autoref{fig2}c). 
\begin{equation}
Eff_{gi}= C_m * E`_{gi}*STE`_{gi}*C_{gi}/(\sum_{1}^{i} E`_{gi}*STE`_{gi}*C_{gi}) \tag{4}\label{eq:4}
\end{equation}
\begin{equation}
Eff_{gi}= Eff_{gi}*RE`_{gi} \tag{5}\label{eq:5}
\end{equation}

After miRNA binds to its target, but might not repress to bound target. The functionality of bound miRNA on target depends on efficiency factors like region that is binding sequence of miRNA on its target. Exact repression efficiency of miRNA is calculated according to equation \eqref{eq:5} (\autoref{fig2}d); $RE’_{gi}$, normalized values of region efficiency coefficient between miRNA and gene. The cooperative repression activity of miRNAs to their common targets is figured out as shown in \autoref{fig2}e.

### Application of model in a breast cancer patient dataset

To determine suitability of method, we have tried to apply model in a real dataset. Firstly, we obtained the miRNA and gene expression datasets of the patient's tumor and normal tissue. We downloaded the high-throughput experimental datasets which are provided miRNA:gene target pairs with interaction factors by **moore_mirnatarget_2015** and [@helwak_mapping_2013]. We have combined miRNA and gene expression datasets via miRNA:target gene dataset **(Whole steps can be found in Supplementary data(TCGA_E9-A1N5_article.Rmd)**). We selected a gene which is over expressed in tumor tissue with compared to normal tissue. It is pointed out that ABCC1 gene, *Multidrug resistance-associated protein 1*, is one of the most significant factor to develop resistance against chemotherapotic agents[@atalay2008role; @lu2015microrna; @atalay2006multidrug]. So, we selected ABCC1 gene for simulation of integrated dataset. We have detected the iteration that is suitable for ABCC1 gene simulation. After simulation of system via ABCC1 gene with over expression, we have compared simulation results and tumor tissue expression levels.

@@

>>>>

---
title: "Network based multifactorial modelling microRNA-target interaction."
author: 
- Selcen Ari^[Department of Bioengineering, Yildiz Technical University, Istanbul, Turkey]
- Alper Yilmaz^[Department of Bioengineering, Yildiz Technical University, Istanbul, Turkey]
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document: 
    keep_tex: true
  html_document: default
bibliography: ref.bib
---

@@

As it is possible for miRNAs to suppress multiple targets, an individual mRNA molecule can also be targeted by multiple miRNAs. In that case, the targeted mRNAs exhibit competitor behavior, that is hypothesized as competing endogenous RNAs (ceRNAs) [@ala_integrated_2013; @cesana_deciphering_2013], against their miRNAs. Briefly, Ala et al. have explained the ceRNA hypothesis as disturbance of the other target when one of the targets on a steady-state system that included one miRNA and two target was perturbed with expression change [@ala_integrated_2013]. Regarding interaction between miRNAs and their target in a cell, explaining and predicting results of an individual perturbation is difficult due to complexity of interactions. Various computational and experimental studies have tackled the problem of unraveling ceRNA:miRNA interactions. It has been observed that when abundance of one of the targets of miR-122 was increased, the other target expression also slightly increased as a result of decreasing repressive activity of miR-122 on them [@denzler_assessing_2014]. Bosson et al. have developed a mathematical model for changes on total target pool concentration after grouping targets according to affinity and demonstrated that miRNA activity correlated with affinity between miRNA and target [@bosson_endogenous_2014]. Cooperative efficiency of miRNAs as well as competitor behaviors of targets were also studied and it has been demonstrated to be crucial for regulating available mRNA levels of targets [@denzler_impact_2016]. MiRNA:target interactions have been modeled as stoichiometric and catalytic mechanisms and Figliuzzi et al. have recommended handling models in network context [@figliuzzi_micrornas_2013]. The model that can explain miRNA target interaction through topological features has been applied at bipartite network by Nitzan et al. [@nitzan_interactions_2014]. Robinson and Henderson applied the model that handles miRNA:target direct and indirect interactions via common miRNA of gene and target of miRNA, at bipartite network. It has been demonstrated that all miRNAs and targets in the network can interact with each other through common miRNAs and genes, without interaction between the same type of nodes [@robinson_modelling_2018].Associated genes that are targets of the same miRNAs have been found with help of correlation of gene expression changes in recent algorithm [@markus_list_sponge_2017]. List et al. have specified that their approach can be useful for ceRNA studies and published their approach as an R package.

## Methods
### Construction of miRNA:target network
miRNA and target pairs per line should be provided as edge list to construct the network. At each line minimum required information is expression levels of miRNA and the target. If available, additional data about factors effecting binding or efficiency of miRNA can be provided as separate columns. After construction of the network, amount of miRNA per target is calculated and kept as edge data. Simply, a target will sequester miRNA proportional to its ratio amount among other targets. If additional criteria effecting the binding of miRNA to its target is provided, distribution of miRNA will be calculated accordingly. Target can be mRNA or any other ceRNA (circRNA, ncRNA, etc.) thus, throughout the manuscript terms target, gene and ceRNA are used interchangeably.

### Triggering perturbation and subsequent calculations
Initially, the network is assumed in steady-state (\autoref{fig1}a) condition and needs a trigger for initiating calculations. The trigger can be a change in expression level of one or more genes (\autoref{fig1}b). After a trigger, the network undergoes iterative cycle of calculations at each of which distribution of miRNA in local neighborhood is recalculated (\autoref{fig1}c). Based on new miRNA distribution, expression level of each node (i.e. ceRNA) is updated. Due to nodes which are targeted by multiple miRNAs, the change in one neighborhood spreads to other neighborhoods (\autoref{fig1}d), consequently have potential to effect whole network due to "ripple effect".

![Schematic presentation of mechanism of network based model. a) In steady state, miRNAs (M: gray triangle) repress targets (G: circle coconut) according to proportion of their targets' expression. b) Two fold increase in transcript level of Gene2 (G2) acts as a trigger. c) Distribution of miRNA1 (M1) changes. d) The change at expression of common target effects changes of proportional distribution of miRNA2 (M2). Expression values are rounded to integers for simplicity](Fig1.jpg){#fig1}

(#TODO caption needs more text, every color should be explained)

During calculations, following assumptions were adopted; 1) Transcription and degradation rates of miRNAs are steady and equal. 2) All available miRNAs are recycled as in miRNA:ceRNA binding, target is degraded and miRNA is unaffected. 3) ceRNA targets also have stable transcription and degradation rates and these rates are equal.

The repression efficiency of a miRNA on the individual target ($Eff_{gi}$) is calculated according to equation \eqref{eq:1}; where miRNA expression ($C_m$) in local neighborhood is distributed among targets using individual gene expression levels ($C_{gi}$) (#TODO isn't $C_g$ gene expressions in the group is equal to total of $C_{gi}$ in group?). For the genes targeted by multiple miRNAs, cooperative activity of miRNAs, $R$, is calculated by summing repression activity of each miRNA (Equation \eqref{eq:2}).

\begin{equation} 
    Eff_{gi}= C_m \times C_{gi}/\sum_{1}^{i} C_g \tag{1}\label{eq:1}
\end{equation}
\begin{equation}
   R_{gi}= Eff_{i1} + Eff_{i2} ... \tag{2}\label{eq:2}
\end{equation}

### Multifactorial calculations in miRNA:target network

Interactions between miRNAs and their targets can be affected from various factors. So, our model integrates multiple factors when calculating overall miRNA activity. We classified factors into two categories. Factors effecting binding determine interaction between miRNA and target and they alter amount of miRNA sequestered to target. Factors effecting efficiency determine proportion of target mRNA degraded by bound miRNA (#TODO fix this sentence). In other words, binding factors exert their influence before or during binding, efficiency factors exert their influence after binding. In the literature, binding free energy [@cao_predicting_2012 ; @helwak_mapping_2013] and seed type [@werfel_preferential_2017] in miRNA:target interactions are described as factors effecting binding affinity. Efficiency factors determine how many of miRNA:target complexes will result in inhibition and binding region on target drastically effect miRNA degradation efficiency [@hausser_analysis_2013; @helwak_mapping_2013]. Both binding and efficiency factors are normalized to their maximum values and scaled to [0,1] interval. The normalized values of factors take into account to determine binding activity and miRNA efficiency on targets (\autoref{fig2}). Binding affinities (activity, $Eff$) of miRNAs on each individual gene are calculated as shown in equation \eqref{eq:4}; $C_m$, miRNA expression in the group; $C_g$, Gene expression; $C_{gi}$, individual gene expression; gi, individual gene; g, whole genes in a group (\autoref{fig2}c). 
\begin{equation}
Eff_{gi}= C_m \times E^\prime_{gi} \times STE^\prime_{gi} \times C_{gi}/(\sum_{1}^{i} E^\prime_{gi} \times STE^\prime_{gi} \times C_{gi}) \tag{4}\label{eq:4}
\end{equation}
\begin{equation}
Eff_{gi}= Eff_{gi}*RE^\prime_{gi} \tag{5}\label{eq:5}
\end{equation}

![Calculations to determine of miRNA binding and repression efficiency. $G$, Gene; $M$, miRNA; $STE$, seed type effect; $RE$, Region Effect; $E$, Energy; $STE^\prime$, normalized values of seed type efficiency coefficient; $RE^\prime$, normalized values of region efficiency coefficient; $E^\prime$, normalized values of energy coefficient.](fig2.jpg){#fig2}

After miRNA binds to its target, but might not repress to bound target. The functionality of bound miRNA on target depends on efficiency factors like region that is binding sequence of miRNA on its target. Exact repression efficiency of miRNA is calculated according to equation \eqref{eq:5} (\autoref{fig2}d); $RE^\prime_{gi}$, normalized values of region efficiency coefficient between miRNA and gene. The cooperative repression activity of miRNAs to their common targets is figured out as shown in \autoref{fig2}e.

### Breast cancer patient dataset

We have applied our model in a real dataset for which experimental measurements of various factors were available. Expression levels of miRNA and genes in tumor and normal tissue of single  patient are retrieved from TCGA(#TODO cite). High-throughput experimental datasets which are provided miRNA:gene target pairs with interaction factors [@helwak_mapping_2013; @moore_mirnatarget_2015]. We have combined miRNA and gene expression datasets via miRNA:target gene dataset retrieved from ... (#TODO name of database and its citation). Detailed description of network construction and its code is available in Supplementary data (#TODO burada link nasıl olacak, word dosyası veya PDF dosyası ismi mi vermemiz mi  gerekiyor, yoksa sadece Sup Data denilmesi yeterli mi? ilgili dosya: TCGA_E9-A1N5_article.Rmd). ABCC1 gene, over-expressed in tumor tissue, was selected as trigger for simulation of integrated dataset. After simulation of network, we have compared simulation results and tumor tissue expression levels.

(#TODO about formulas, 1) eq:3 F/Fmax can be removed, 2) in formulas eq:1 and eq:2, C_m or Eff_gi is not intuitive, is there consensus terms about effect or nth gene? 3) in formula eq:4, instead of, 4) in general, `*` sign should be replaced with `\times` and apostrophe should be replaced with `\prime` )

@@

(#TODO BRCA results go here) ABCC1 gene, *Multidrug resistance-associated protein 1*, is one of the most significant factor to develop resistance against chemotherapotic agents[@atalay2008role; @lu2015microrna; @atalay2006multidrug]
